<!-- README.md is generated from README.Rmd. Edit README.Rmd, then run
     knitr::knit("README.Rmd") with the installed package. -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 80, scipen = 999)
```

# mediaPlanR

[![R-CMD-check](https://github.com/majesus/mediaPlanR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/majesus/mediaPlanR/actions/workflows/R-CMD-check.yaml)

**mediaPlanR estimates how many people a media plan reaches, and how many
times, from the audience figures a planner already has.**

It brings together, in one package, the classical reach and
exposure-distribution models of the media-planning literature (Sainsbury,
Binomial, Beta-Binomial, Metheringham, Agostini, Hofmans, CANEX, CBD, CSD, MSAD
and MBD) and the tools around them: a validated plan object, plan metrics, model
comparison, target-audience metrics, comparison with observed data, and
budget allocation that states whether the optimum it found was verified.

The package is written for students and practitioners of media planning. This
document is meant to be read in order: first the ideas, then a complete plan
worked from start to finish, then how to choose a model and how to use your own
data. The outputs shown below were produced by the code next to them.

## Installation

```r
# install.packages("pak")
pak::pak("majesus/mediaPlanR")
```

mediaPlanR requires R 4.0 or later. It contains no compiled code, but on
Windows `pak` may stop with *Could not find tools necessary to compile a
package* when it builds from source. You can either install
[Rtools](https://cran.r-project.org/bin/windows/Rtools/) (the version that
matches your R) or install the package without building it:

```r
install.packages("remotes")
remotes::install_github("majesus/mediaPlanR", build = FALSE)
```

This second route skips the vignette. Then load the package:

```{r}
library(mediaPlanR)
```

## The ideas in five minutes

A media plan buys *insertions* (an advertisement in an issue, a spot, a post) in
*vehicles* (a newspaper, a radio station, a website). The planner knows how many
people each insertion exposes, the **audience**, and wants to know how many
different people the plan exposes and how often. These terms are used throughout:

| Term | Meaning |
|---|---|
| **Impressions** | Total exposures: audience times insertions, added over the plan. A person exposed twice counts twice. |
| **GRP** (gross rating points) | Impressions as a percentage of the population. |
| **Reach** | The share of the population exposed at least once. Each person counts once. |
| **Exposure distribution** | The share of the population exposed exactly 0, 1, 2, ... times. |
| **Average frequency** | Average number of exposures among the people reached: impressions divided by people reached. |
| **Effective reach** | The share exposed at least *k* times, where *k* (the effective frequency) is chosen by the planner. |
| **Duplication** | The people who belong to the audience of two vehicles. It is the reason reach is smaller than the sum of the audiences. |

Take a population of 1,000,000 people and two vehicles with one insertion each,
reaching 300,000 and 400,000 people:

```{r}
population <- 1000000
audiences  <- c(A = 300000, B = 400000)

sum(audiences)                       # impressions
sum(audiences) / population * 100    # GRP
```

The two audiences add up to 700,000 impressions, but fewer than 700,000 *different*
people are reached, because some are in both audiences. If the two audiences
overlap only by chance (**random duplication**), the share of the population in
both is the product of the two shares, and the reach follows directly:

```{r}
duplicated_people <- prod(audiences / population) * population
duplicated_people                           # people expected in both audiences
sum(audiences) - duplicated_people          # reach, in people
```

`calc_sainsbury()` does this calculation and adds the full exposure
distribution:

```{r}
calc_sainsbury(audiences, population)
```

The result can be read by hand: 42% of the population (0.7 x 0.6) is in
neither audience, 46% is in exactly one and 12% (0.3 x 0.4) is in both. Reach
is therefore 58%, and the average frequency among the people reached is
700,000 / 580,000 = 1.21.

In practice duplication is *observed*, for example in an audience survey. If
150,000 people are in both audiences, reach is 300,000 + 400,000 - 150,000 =
550,000 people (55%), lower than under random duplication. Most models in the
package exist to use such observed duplication when the plan has several
vehicles and several insertions, where no simple formula is available.

## A first plan, from start to finish

`media_plan()` stores a plan in explicit units: audience in people per
insertion, cost in currency units per insertion and the population in people.
It rejects inconsistent inputs (for example an audience larger than the
population) instead of calculating with them. The optional `target_audience`
column holds the people of your target group in each insertion.

```{r}
plan <- media_plan(
  data.frame(
    channel = c("TV", "Radio", "Digital"),
    audience = c(250000, 150000, 100000),
    insertions = c(3, 4, 6),
    cost_per_insertion = c(15000, 3000, 800),
    target_audience = c(125000, 75000, 60000)
  ),
  population = 1000000,
  target_audience = "target_audience"
)
plan
```

`plan_metrics()` gives impressions, spend, GRP and cost ratios, channel by
channel and for the whole plan. Every ratio has an explicit denominator:

```{r}
metrics <- plan_metrics(plan)
metrics$by_channel[, c("channel", "impressions", "spend", "rating_points",
                       "cpm_impressions")]
metrics$totals[c("impressions", "spend", "grps")]
```

`estimate_reach()` estimates reach and the exposure distribution. Its default
model is Sainsbury's, which assumes random duplication:

```{r}
reach <- estimate_reach(plan, model = "sainsbury")
reach
```

The result is a list. Reach is available in probability, percent and people,
and the exposure distribution includes the zero-exposure cell, so it always sums
to one:

```{r}
reach$reach$percent
round(reach$distribution[, c("contacts", "percent")], 2)
sum(reach$distribution$probability)
```

The `cumulative` element gives the share exposed *at least* N times. The row for
the effective frequency you choose is the effective reach: with an effective
frequency of 2, it is the 60.5% shown for `min_contacts = 2`.

```{r}
reach$cumulative[reach$cumulative$min_contacts %in% 1:4,
                 c("min_contacts", "percent", "people")]
```

Passing the estimated reach back to `plan_metrics()` adds the metrics that need
it, such as the cost of reaching a thousand different people:

```{r}
plan_metrics(plan, reach = reach$reach$people)$totals[
  c("reach_percent", "average_frequency", "cost_per_thousand_reached")]
```

`compare_reach_models()` places the two models that need no duplication data
side by side. Sainsbury lets each vehicle keep its own audience; Binomial gives
every insertion the plan's insertion-weighted average probability:

```{r}
compare_reach_models(plan, c("sainsbury", "binomial"))
```

### Allocating a budget

`optimize_media_plan()` searches for the number of insertions in each channel
that maximizes effective reach within a budget (`objective = "max_reach"`) or
that reaches a required effective reach at the lowest cost
(`objective = "min_cost"`). Here, effective reach means being exposed at least
twice, and `max_insertions` caps the insertions of each channel (the search is
not limited to the insertions of the original plan):

```{r}
optimized <- optimize_media_plan(
  plan,
  budget = 50000,
  objective = "max_reach",
  effective_frequency = 2,
  max_insertions = c(4, 8, 12)
)
optimized
optimized$allocation
optimized$global_optimum
```

When the number of possible allocations is manageable, the function evaluates
all of them and `global_optimum` is `TRUE`: the allocation is the best one for
the model used. Otherwise it uses a greedy search, labels the result as a
heuristic and never reports it as a global optimum.

### Target-audience metrics

A channel with a large audience is not necessarily a good channel for your
target. `audience_metrics()` separates three quantities that are often mixed:
the share of a channel's audience that belongs to the target (*composition*),
the share of the target that the channel reaches (*target rating*) and the
*affinity index*, which compares the target's share of the channel's audience
with its share of the population (100 means the channel is neutral):

```{r}
audience_metrics(
  gross_audience = c(250000, 150000),
  target_audience = c(125000, 75000),
  gross_universe = 1000000,
  target_universe = 400000
)
```

## Why the choice of model matters

Sainsbury and Binomial assume that every insertion is an independent
opportunity, even when two insertions are in the same vehicle. For different
vehicles that is the random-duplication hypothesis of the previous section. For
repeated insertions in one vehicle it is a strong assumption, because the people
who see the first insertion tend to be the same people who see the next one.

Consider one vehicle with a 500,000-person audience in a population of
1,000,000, and five insertions. Under random accumulation each insertion is an
independent 50% chance, so almost everybody is reached:

```{r}
calc_binomial(audiences = 500000, population = 1000000, insertions = 5)$reach$percent
```

Now suppose that audience research shows that the cumulative audience after two
insertions is only 550,000: the second insertion adds just 50,000 new people.
That is evidence of a loyal audience, and `calc_beta_binomial()` uses it. The
**Beta-Binomial** model lets every person have a different, personal probability
of exposure and estimates how those probabilities are spread from the two
audience figures:

```{r}
beta_binomial <- calc_beta_binomial(A1 = 500000, A2 = 550000, P = 1000000, n = 5)
beta_binomial$reach$percent
beta_binomial$distribution$percent
beta_binomial$parameters$zero_contact_probability
```

Reach falls from 96.9% to 60.3%, and the distribution is polarized: 39.7% of the
population is exposed all five times (the last element of the distribution) and
39.7% is never exposed (the zero-exposure percentage shown last). These numbers are illustrative, but the lesson is general: **the same
vehicles and insertions can give very different reach depending on how
audiences repeat and overlap**, and the model is how that behavior enters the
estimate. The next section explains how to choose.

## Choosing a model

The models are organized by the shape of plan they were derived for (Aldás
Manzano, 1998) and by the data they need.

| Plan | Data you need | Models |
|---|---|---|
| Several vehicles, any number of insertions, no duplication data | Audience of each vehicle, population | `calc_sainsbury()`, `calc_binomial()` |
| One vehicle, several insertions | Audience after one and after two insertions | `calc_beta_binomial()`, `calc_hofmans_accumulation()` |
| Several vehicles, one insertion each, with observed duplication | Audiences and the duplication of each pair | `calc_agostini_duplication()`, `calc_hofmans_duplication()` (reach only) |
| Several vehicles and insertions, with observed duplication | Reach after one and two insertions of each vehicle, and the duplication of each pair | `calc_metheringham()`, `calc_canex()`, `calc_cbd()`, `calc_csd()`, `calc_msad()`, `calc_mbd()` |
| One Beta-Binomial for a whole schedule, matched to a reach figure you already have | Insertions, audiences and that reach | `fit_bbd_to_reach()`; `calibrate_bbd()` matches a target effective reach instead |
| Counts that are not limited by a number of insertions (page views, ad-server exposures) | Observed counts, or a mean and a dispersion | `fit_nbd_exposure()`, `nbd_exposure_distribution()` |

Some guidance for the common situations:

- **You only have audience figures.** Use Sainsbury or Binomial. They are exact
  under their assumptions, but those assumptions are strong, as the previous
  section showed: when audiences repeat or overlap more than chance would, they
  overstate reach.
- **You have a vehicle's audience after one and two insertions.** Use the
  Beta-Binomial model for that vehicle.
- **You have observed duplication between vehicles.** For one insertion per
  vehicle, the Agostini and Hofmans formulas return reach only. For several
  insertions, Metheringham's model is the simplest: it replaces the plan by an
  "average vehicle". CANEX, CBD, CSD, MSAD and MBD model each vehicle separately
  and combine them in different ways; they are more demanding and their
  differences are described in `vignette("mediaPlanR-intro")`.
- **Your exposures are unbounded counts.** The Negative-Binomial functions are
  scoped to those processes. They are deliberately not offered for finite
  insertion schedules, and they are not available in `estimate_reach()` or
  `optimize_media_plan()`.

## Using your own data

Every model that needs more than audiences takes a set of arguments that mirror
a small table. The quickest way to see what a function expects is its example
dataset, which is already a list of arguments:

```{r}
data(csd_kim2005)     # reach and duplication figures of a worked example, Kim (2005)
str(csd_kim2005)
```

so that `do.call()` runs the model on it. Printing a result gives the full report
of the model: headline metrics, parameters, the exposure distribution, the
cumulative distribution and diagnostics.

```{r}
csd <- do.call(calc_csd, csd_kim2005)
csd
```

Example datasets exist for the main models (`ratings_example`,
`beta_binomial_example`, `hofmans_accumulation_example`, `duplication_example`,
`metheringham_example`, `canex_example`, `csd_example`, `msad_example`,
`mbd_example` and `bbd_reach_example`). `csd_kim2005`, `msad_kim2005` and
`mbd_cheong2007` reproduce the published figures of the worked examples of Kim
(2005) and Cheong (2007) so that you can check the package against the sources.

### CSD, MSAD, CBD and MBD

These four models take the same three kinds of input, all as **proportions of
the population** (between 0 and 1):

- `vehicles_data`: one row per vehicle with `insertions` (planned insertions, a
  whole number), `R1` (reach after one insertion, that is, the vehicle's
  audience) and `R2` (cumulative reach after two insertions). `R2` must satisfy
  `R1 <= R2 <= 2 * R1 - R1^2`, where the upper limit is random accumulation. Use
  `NA` for vehicles with a single insertion. Extra columns, such as a vehicle
  name, must be left out.
- `duplications`: a symmetric square matrix with the proportion of the
  population reached by both vehicles with one insertion each. The diagonal is
  ignored. Rows and columns follow the order of `vehicles_data`. A duplication
  cannot exceed the smaller of the two audiences.
- `population`: the number of people in the population, so that results are also
  expressed in people. The default, 1, leaves them as proportions.

```{r}
vehicles <- data.frame(
  vehicle    = c("TV", "Radio", "Digital"),
  insertions = c(3, 2, 4),
  R1         = c(0.35, 0.18, 0.10),
  R2         = c(0.44, 0.24, 0.15)
)

duplications <- matrix(
  c(NA,   0.08, 0.05,
    0.08, NA,   0.03,
    0.05, 0.03, NA),
  nrow = 3, byrow = TRUE,
  dimnames = list(vehicles$vehicle, vehicles$vehicle)
)

csd <- calc_csd(
  vehicles_data = vehicles[, c("insertions", "R1", "R2")],
  duplications = duplications,
  aggregation_order = "audience_desc",
  population = 8000000
)
csd
```

The report of these four models also lists the vehicles and the aggregation
steps. Their distributions have one row per insertion of the plan, so a long
one is cut after `max_rows` exposure levels (30 by default; use
`print(csd, max_rows = Inf)` to list them all). For a four-line summary, use
`print(csd, full = FALSE)`.

`aggregation_order` can be `"audience_desc"` (largest audience first), `"given"`
(the row order of `vehicles_data`) or a permutation such as `c(2, 1, 3)`. The
order can change the exposure distribution of CSD, MSAD and MBD, so report the one
you used (CBD does not depend on it). The other three models take the same
arguments; they use different formulas and therefore give different results
from the same inputs (CSD and CBD obtain reach from the same canonical
expansion, which is why their reach coincides here):

```{r}
v <- vehicles[, c("insertions", "R1", "R2")]
models <- list(CSD = calc_csd, MSAD = calc_msad, CBD = calc_cbd, MBD = calc_mbd)
data.frame(
  model = names(models),
  reach_percent = sapply(models, function(f) f(v, duplications)$reach$percent),
  average_frequency = sapply(models, function(f) f(v, duplications)$average_frequency),
  row.names = NULL
)
```

The models do not accept every input. When the duplications are incompatible
with the model, the function stops with a message that explains why; it does
not silently alter your data. For example, `calc_mbd()` rejects the same plan
with duplications of 0.07, 0.04 and 0.02, because together they imply less
overlap than random duplication would for the Beta-Binomial distribution that
the model imputes for three vehicles.

### The other models

```{r}
# Random duplication: three vehicles with 2, 1 and 3 insertions
calc_sainsbury(audiences = c(300000, 400000, 200000),
               population = 1000000, insertions = c(2, 1, 3))$reach$percent

# Several vehicles, one insertion each: duplication matrix in people
duplication <- matrix(c(NA,     140000, 70000,
                        140000, NA,     90000,
                        70000,  90000,  NA), nrow = 3, byrow = TRUE)
calc_hofmans_duplication(audiences = c(300000, 400000, 200000),
                         population = 1000000,
                         duplication_matrix = duplication)$reach$percent

# Several vehicles and insertions, duplication observed in people. The diagonal
# holds the duplication between two insertions of the same vehicle, so it is
# not NA.
calc_metheringham(
  audiences = c(1500000, 800000, 1200000),
  insertions = c(4, 3, 5),
  duplication_matrix = matrix(c(150000, 200000, 180000,
                                200000, 120000, 140000,
                                180000, 140000, 170000), nrow = 3),
  population = 10000000
)$reach$percent
```

Units differ between families, and the functions check them:

| Model | Function | Main inputs | Units |
|---|---|---|---|
| Sainsbury, Binomial | `calc_sainsbury()`, `calc_binomial()` | `audiences`, `population`, `insertions` | people |
| Beta-Binomial | `calc_beta_binomial()` | `A1`, `A2`, `P`, `n` | people |
| Hofmans (accumulation) | `calc_hofmans_accumulation()` | `R1`, `R2`, `N` | proportions |
| Agostini, Hofmans (duplication) | `calc_agostini_duplication()`, `calc_hofmans_duplication()` | `audiences`, `population`, `duplication_matrix` | people |
| Metheringham | `calc_metheringham()` | `audiences`, `insertions`, `duplication_matrix`, `population` | people |
| CANEX | `calc_canex()` | `vehicles_data` (`k`, `R1`, `R2`), `duplications`, `population` | proportions |
| CSD, MSAD, CBD, MBD | `calc_csd()`, `calc_msad()`, `calc_cbd()`, `calc_mbd()` | `vehicles_data` (`insertions`, `R1`, `R2`), `duplications`, `population` | proportions |

Each function's help page (`?calc_canex`, `?calc_cbd`, ...) documents its
arguments, its assumptions and the published sources.

## Reading the results

Results have a common core, but their shape depends on the family of models:

| Result | Reach | Exposure distribution |
|---|---|---|
| `estimate_reach()` | `$reach`: `probability`, `percent`, `people` | `$distribution`: data frame, from 0 exposures |
| `calc_sainsbury()`, `calc_binomial()`, `calc_beta_binomial()`, `calc_metheringham()` | `$reach`: `percent`, `people` | `$distribution`: list of `percent` and `people`, from 1 exposure; the share with no exposure is 100 minus the reach |
| `calc_canex()`, `calc_csd()`, `calc_msad()`, `calc_cbd()`, `calc_mbd()` | `$reach`: `probability`, `percent`, `people`; also `$average_frequency` | `$distribution`: data frame, from 0 exposures |
| `calc_agostini_duplication()`, `calc_hofmans_duplication()` | `$reach` only | not estimated |

When you calculate with the objects yourself, two quick coherence checks are
that the probabilities add up to one and that the average frequency equals the
mean number of exposures per person divided by the reach.

```{r}
sum(csd$distribution$probability)
csd$average_frequency
sum(vehicles$insertions * vehicles$R1) / csd$reach$probability
```

The last two numbers coincide because CSD preserves the plan's mean number of
exposures per person (the sum of insertions times first-insertion reach). MSAD
and MBD do too; CBD reproduces it only approximately, and reports the difference
in `diagnostics$mean_error`.

## Comparing a model with observed data

When you have a measured exposure distribution for a schedule, for example from
a panel, `evaluate_exposure_model()` compares it with a model. The table must
contain `contacts` and `observed`, include the zero-exposure cell, and declare
its scale (`"count"`, `"probability"` or `"percent"`). The counts below are
hypothetical: a panel of 1,000 people classified by the number of exposures to
the three-vehicle schedule of the previous sections (0 to 9 exposures).

```{r}
observed <- data.frame(
  contacts = 0:9,
  observed = c(330, 150, 160, 190, 80, 50, 25, 10, 5, 0)
)

evaluation <- evaluate_exposure_model(
  observed = observed,
  predicted = csd,
  observed_scale = "count"
)
round(unlist(evaluation$summary[c("kim_aer", "kim_ape")]), 3)
```

`kim_aer` is the average relative error in reach, |observed - predicted| /
observed. `kim_ape` is the average relative error in the exposure distribution:
the absolute differences over the exposure levels of one or more, divided by the
observed reach (both are proposed by Kim, 2005). `evaluation$summary` also holds
the total variation distance and other diagnostics. These errors are descriptive
measures of fit, not statistical tests. The function
requires the observed and predicted distributions to cover exactly the same
exposure levels. Observed data are always supplied by the analyst; the package
never infers them from model inputs.

## Limits to keep in mind

- **Audience is an opportunity to see, not attention.** The models estimate how
  many people are exposed to a vehicle, not how many notice, remember or are
  persuaded by the message.
- **Every model is an approximation.** Each one rests on stated assumptions about
  duplication and about how exposure probabilities vary between people. Where the
  assumption is wrong, so is the estimate, and the help pages state which
  assumption each model makes.
- **Observed inputs must be coherent.** The multivariate models stop when
  pairwise duplications are impossible given the audiences, or when some three
  vehicles have no compatible joint exposure probabilities. CANEX can produce small
  negative probabilities; it sets them to zero and reports in `diagnostics`
  how much probability was altered. MBD can do so too when it combines four or
  more vehicles, and warns when this is possible.
- **Size is limited.** The number of exposure combinations grows exponentially
  with vehicles and insertions. CANEX stops with an error on plans that are too
  large for it, and MBD is limited to 12 vehicles.
- **Published values are reproduced up to documented differences.** Where the
  package differs from a published table, `vignette("mediaPlanR-intro")` and the
  file `system.file("DATA-PROVENANCE.md", package = "mediaPlanR")` state by how
  much and why.

## Learn more

- `vignette("mediaPlanR-intro")` develops each model, its assumptions and its
  validation against the sources.
- Each function has a help page with its arguments, details, references and
  examples: start with `?media_plan`, `?estimate_reach` and `?calc_csd`.
- `citation("mediaPlanR")` gives the reference for the package. When you use a
  specific model, please also cite the original source of that model.
- Questions, errors and suggestions are welcome in the
  [issue tracker](https://github.com/majesus/mediaPlanR/issues).

## References

- Agostini, J. M. (1961). How to estimate unduplicated audiences. *Journal of
  Advertising Research*, 1(3), 11-14. <https://doi.org/10.1080/00218499.1961.12519620>
- Aldás Manzano, J. (1998). *Modelos de determinación de la cobertura y la
  distribución de contactos en la planificación de medios publicitarios
  impresos*. Doctoral dissertation, Universidad de Valencia.
- Cheong, Y. (2007). *Multivariate Beta Binomial Distribution Model as a Web
  Media Exposure Model*. Doctoral dissertation, The University of Texas at
  Austin.
- Cheong, Y., Leckenby, J. D., & Eakin, T. (2011). Evaluating the multivariate
  beta binomial distribution for estimating magazine and Internet exposure
  frequency distributions. *Journal of Advertising*, 40(1), 7-23.
  <https://doi.org/10.2753/JOA0091-3367400101>
- Danaher, P. J. (1991). A canonical expansion model for multivariate media
  exposure distributions: A generalization of the "duplication of viewing law".
  *Journal of Marketing Research*, 28(3), 361-367.
  <https://doi.org/10.1177/002224379102800311>
- Hofmans, P. (1966). Measuring the cumulative net coverage of any combination of
  media. *Journal of Marketing Research*, 3(3), 269-278.
  <https://doi.org/10.1177/002224376600300307>
- Kim, H. (1994). *A conditional beta distribution model for advertising media
  reach/frequency estimation*. Doctoral dissertation, The University of Texas at
  Austin.
- Kim, H. G. (2005). *A Canonical Sequential Aggregation Media Model*. Doctoral
  dissertation, The University of Texas at Austin.
- Leckenby, J. D., & Rice, M. D. (1986). The declining reach phenomenon in
  exposure distribution models. *Journal of Advertising*, 15(3), 13-20.
  <https://doi.org/10.1080/00913367.1986.10673014>
- Metheringham, R. A. (1964). Measuring the net cumulative coverage of a print
  campaign. *Journal of Advertising Research*, 4(4), 23-28.
  <https://doi.org/10.1080/00218499.1964.12519751>

## Author and license

Manuel J. Sánchez-Franco, Universidad de Sevilla.
[ORCID 0000-0002-8042-3550](https://orcid.org/0000-0002-8042-3550)

Released under the MIT license.
