---
title: "Introduction to mediaPlanR"
author: "Manuel J. Sánchez-Franco"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to mediaPlanR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## What mediaPlanR provides

**mediaPlanR** brings together, in a single R package, the classical reach
and exposure-distribution models used in advertising media planning, along
with a validated planning core (plan, metrics, model comparison, and
budget-constrained reach optimization).

```{r setup}
library(mediaPlanR)
```

## The version 2 workflow

The main API starts from a validated plan and preserves each variable's
units. Audience represents people per insertion; cost represents currency
per insertion; population is the planning universe.

```{r v2-plan}
plan <- media_plan(
  data.frame(
    channel = c("TV", "Radio", "Digital"),
    audience = c(300000, 180000, 120000),
    insertions = c(4, 6, 10),
    cost_per_insertion = c(18000, 3500, 1200),
    target_audience = c(180000, 90000, 84000)
  ),
  population = 1000000,
  target_audience = "target_audience"
)

metrics <- plan_metrics(plan)
reach <- estimate_reach(plan, model = "sainsbury")
reach
```

The returned distribution includes zero exposures. This lets you check that
the probabilities sum to one, that reach equals `1 - P(X=0)`, and that
average frequency matches impressions divided by people reached.

```{r v2-invariants}
sum(reach$distribution$probability)
reach$reach$probability
reach$average_frequency
```

`estimate_reach()` is a `media_plan`-native front end for
`calc_sainsbury()`/`calc_binomial()`: it reads `audience`, `insertions`, and
`population` from the plan and calls the requested model directly, so the
two are backed by one implementation, not two. `model = "sainsbury"` keeps
each vehicle's own audience (heterogeneous); `model = "binomial"` instead
uses the plan's average probability for every vehicle (homogeneous). Both
assume random duplication *and* random accumulation: a repeat insertion in
the same vehicle is treated as an independent trial, exactly like an
insertion in a different vehicle. That is a real simplification, not a bug
-- it is the hypothesis the cited models state explicitly (Lee & Burkart,
1960; Chandon, 1985 -- reviewed in Aldas Manzano, 1998, secs.
3.3.1.1-3.3.1.2), and it is what lets them run without any observed
duplication data. With one insertion per vehicle, both reduce exactly to
their respective one-insertion-per-vehicle formulation (Aldas Manzano, 1998,
secs. 3.2.2.1-3.2.2.2).

### Optimization with an optimality diagnostic

```{r v2-optimization}
optimized <- optimize_media_plan(
  plan,
  budget = 60000,
  objective = "max_reach",
  effective_frequency = 2,
  max_insertions = c(4, 8, 12),
  method = "auto"
)

optimized$allocation
optimized$global_optimum
```

When the search space is manageable it is evaluated exhaustively and
`global_optimum = TRUE`. In large spaces, the heuristic mode is identified as
such; it is never presented as a global optimum. Candidate allocations
routinely place several insertions in the same vehicle, so the optimizer
only offers `sainsbury`/`binomial` as its reach model -- the experimental NBD
approximation is scoped to continuous exposure processes, not finite
insertion schedules, and is not offered here.

## The family of reach and exposure-distribution models

All of these models answer the same question -- *what proportion of the
target population is reached, and how many times, by a given media plan?*
-- but they differ in the **shape of plan** they were derived for, and in
the assumptions they make about the **duplication** of audiences across
vehicles and about the **heterogeneity** of individual exposure
probabilities. Plan shape follows Aldas Manzano's (1998) own three-way
split, which this package keeps deliberately visible: Beta-Binomial and
Hofmans (accumulation) are derived for one vehicle and do not generalize to
several, and Agostini (duplication) and Hofmans (duplication) are derived
for several vehicles with one insertion each and do not generalize to
several insertions. Only Sainsbury and Binomial cover both cases, because
each is the same model at two levels of generality (see `insertions` above).

| Model | Function | Plan shape | Assumed duplication | Heterogeneity |
|---|---|---|---|---|
| Sainsbury | `calc_sainsbury()` / `estimate_reach(model = "sainsbury")` | 1 or more insertions, m vehicles | Random (independence and accumulation) | Yes, across vehicles |
| Binomial | `calc_binomial()` / `estimate_reach(model = "binomial")` | 1 or more insertions, m vehicles | Random (independence and accumulation) | No (average vehicle) |
| Agostini (duplication) | `calc_agostini_duplication()` | 1 insertion, m vehicles | Corrected with a single k coefficient | Random + empirical correction |
| Hofmans (duplication) | `calc_hofmans_duplication()` | 1 insertion, m vehicles | Corrected with a pairwise coefficient per vehicle pair | Empirical, no distribution |
| Beta-Binomial | `calc_beta_binomial()` | 1 vehicle, n insertions | Implicit (one vehicle, n insertions) | Yes, across individuals (Binomial-Beta mixture) |
| Hofmans (accumulation) | `calc_hofmans_accumulation()` | 1 vehicle, n insertions | Implicit (curve fitted to R1, R2) | Non-linear empirical curve |
| Metheringham | `calc_metheringham()` | n insertions, m vehicles | Observed (averaged) | Yes, via a Beta distribution |
| CANEX | `calc_canex()` | n insertions, m vehicles | Observed, correlation matrix | Yes, per vehicle (Beta-Binomial) |
| CBD (conditional Beta) | `calc_cbd()` | n insertions, m vehicles | Observed; canonical expansion at the (0,1) level | Yes, per-vehicle BBD marginal |
| BBD fitted to external reach | `fit_bbd_to_reach()` | n insertions, m vehicles | Reach is supplied as a constraint | Yes, via Beta-Binomial |
| CSD (canonical sequential aggregation) | `calc_csd()` | n insertions, m vehicles | Observed; canonical reach per sub-schedule with preserved margins | Yes, per-vehicle BBD marginal |
| MSAD (Morgensztern) | `calc_msad()` | n insertions, m vehicles | Observed and preserved during sequential aggregation | Yes, per-vehicle BBD marginal |
| MBD (multivariate Beta-Binomial) | `calc_mbd()` | n insertions, m vehicles | Observed pairs; imputed for 3+ vehicles via Waring's theorem | Yes, per-vehicle BBD marginal |
| Experimental NBD | `nbd_exposure_distribution()` | Continuous process (not a finite schedule) | Does not represent cross-vehicle dependence | Yes, Poisson-Gamma heterogeneity in an unbounded process |

### Example: three vehicles, the same plan, different assumptions

```{r}
audiences <- c(300000, 400000, 200000)
population <- 1000000

reach_sainsbury <- calc_sainsbury(audiences, population)
reach_binomial <- calc_binomial(audiences, population)
reach_agostini_independence <- calc_agostini_duplication(audiences, population, k = 1)
reach_agostini_corrected <- calc_agostini_duplication(audiences, population, k = 0.9)

data(hofmans_duplication_example)
reach_hofmans_duplication <- do.call(calc_hofmans_duplication, hofmans_duplication_example)

data.frame(
  model = c("Sainsbury", "Binomial", "Agostini (k=1)", "Agostini (k=0.9)",
            "Hofmans (duplication)"),
  reach_pct = round(c(
    reach_sainsbury$reach$percent,
    reach_binomial$reach$percent,
    reach_agostini_independence$reach$percent,
    reach_agostini_corrected$reach$percent,
    reach_hofmans_duplication$reach$percent
  ), 2)
)
```

Note that, with `k = 1`, `calc_agostini_duplication()` matches
`calc_sainsbury()` exactly: both start from the same random-duplication
assumption, applied iteratively in Agostini's case. With `k = 0.9`,
predicted duplication is reduced and estimated reach therefore increases
relative to the pure independence assumption. `calc_hofmans_duplication()`
uses observed pairwise duplication instead of an assumed or fitted
coefficient, so its reach need not fall between the other four -- here the
observed overlap is lower than random duplication would predict, so its
reach is the highest of the group.

## Beta-Binomial vs. NBD: finite opportunities and count processes

Both `calc_beta_binomial()` and the NBD assume each individual has their own
personal exposure probability (or rate), heterogeneous across the
population, but they differ in their distributional assumption: the former
mixes a Binomial (n fixed insertions) with a Beta; the NBD mixes a Poisson
with a Gamma. That difference is substantive: the NBD has unbounded support
and is appropriate for page views, ad-server exposures, or other processes
with continuous opportunities -- it should not be used to imply that a
traditional, finite-insertion schedule can register more exposures than
insertions.

When individual observed counts are available, `fit_nbd_exposure()`
estimates the mean and heterogeneity by maximum likelihood. For scenarios
with external parameters, `nbd_exposure_distribution()` keeps the last cell
as an open tail and never presents it as an exact count:

```{r}
counts <- c(rep(0, 40), rep(1, 25), rep(2, 15), rep(3, 8), 5, 7)
nbd_fit <- fit_nbd_exposure(counts)
nbd_fit$estimates

nbd_scenario <- nbd_exposure_distribution(
  mean_contacts = 2.5, size = 1.7, report_max = 9, opportunities = 9
)
nbd_scenario$diagnostics$probability_above_opportunities
```

That last diagnostic quantifies the mass an unbounded NBD assigns above the
finite opportunities.

## CSD and MSAD: two reach estimators, one sequential architecture

`calc_csd()` reproduces Kim's (2005) complete numerical example: three
vehicles, BBD marginals, second-order canonical-expansion reach, and
non-random convolution with margin preservation.

```{r}
data(csd_kim2005)
csd_result <- do.call(calc_csd, csd_kim2005)
csd_result$reach
csd_result$distribution
csd_result$steps
```

The exact results may differ by a few hundredths of a percentage point from
Kim's tables because the thesis rounds parameters and intermediate matrices
to four decimals. The function preserves the precision of its inputs.

### MSAD: Morgensztern's sequential aggregation

`calc_msad()` implements the structure described by Leckenby and Rice (1986)
and Kim (2005): a per-vehicle BBD marginal, Morgensztern reach for each
sub-schedule, and non-random convolution that preserves the marginals and
sets the zero-exposure probability to `1 - reach`.

```{r}
data(msad_kim2005)
msad_result <- do.call(calc_msad, msad_kim2005)
msad_result$reach
msad_result$steps
msad_result$diagnostics
```

The default rule aggregates the highest-audience vehicles first, as Kim
indicates. An explicit permutation can also be supplied to test sensitivity
to order. The function stops if the duplications or the Morgensztern reach
violate probabilistic bounds.

Kim specifies MSAD but refers to Lee (1988) for the complete numerical
example. `msad_kim2005` therefore reuses the inputs from Kim's CSD example
and labels its output as a derived calculation, not an MSAD result published
in the thesis.

## Validation against observed data

Observed data are always supplied by the analyst. For one schedule, they
must be provided as a table with the columns `contacts` and `observed`,
explicitly including zero exposures. Whether the values are counts,
probabilities, or percentages must also be declared:

```{r}
kim_observed <- data.frame(
  contacts = 0:6,
  observed = c(38.41, 17.89, 39.66, 2.67, 1.36, 0, 0)
)

csd_evaluation <- evaluate_exposure_model(
  observed = kim_observed,
  predicted = csd_result,
  observed_scale = "percent"
)

csd_evaluation$summary
csd_evaluation$by_schedule
```

If the prediction is also supplied as a table, its columns must be named
`contacts` and `predicted`, and `predicted_scale` must be declared. To
evaluate several schedules at once, add an identifier column to both tables
and name it via `schedule_col`.

The evaluator requires identical supports. It does not interpret an NBD open
tail as an exact frequency: before evaluation, the observed and predicted
tails must be collapsed at the same threshold. Besides Kim's historical
AER/APE, it also returns absolute reach error, total variation, cell MAE,
and mean-exposure bias.

## The CANEX model: heterogeneity and jointly observed duplication

`calc_canex()` implements Danaher's (1991) canonical expansion model, which
combines a per-vehicle Beta-Binomial distribution with a correlation term
that adjusts the joint probability according to the observed duplication
between each pair of vehicles:

```{r}
vehicles <- data.frame(
  k = c(2, 2),
  R1 = c(0.4902, 0.033),
  R2 = c(0.5805, 0.0502)
)
duplications <- matrix(
  c(1.000, 0.0157,
    0.0157, 1.000),
  nrow = 2, byrow = TRUE
)

canex_result <- calc_canex(vehicles, duplications, population = 1000000)
canex_result
```

The number of exposure combinations to evaluate grows exponentially with the
number of vehicles and insertions, so the model is intended for small- or
medium-sized schedules; for very large schedules, `calc_canex()` stops with
an informative error instead of exhausting available memory.

## Ready-to-use example datasets

Models with more complex input structures have datasets with descriptive
names, whose elements match the function's arguments:

```{r}
data(canex_example)
str(canex_example)
do.call(calc_canex, canex_example)$total_reach
```

For CSD, MSAD, and MBD there are two example datasets per model.
`csd_example`, `msad_example`, and `mbd_example` are original, illustrative,
self-contained scenarios meant for quick experimentation with `do.call()`.
`csd_kim2005`, `msad_kim2005`, and `mbd_cheong2007` instead reproduce the
minimal numeric values (audiences and duplication figures) published by Kim
(2005) and Cheong (2007) in their doctoral dissertations, included solely so
anyone can verify that `calc_csd()`, `calc_msad()`, and `calc_mbd()`
reproduce their respective published examples: they are bare factual data
points reused for validation, not a creative or substantial reproduction of
those dissertations. `bbd_reach_example` plays the same illustrative role
for `fit_bbd_to_reach()`.
`calc_sainsbury()`, `calc_binomial()`, and `calc_agostini_duplication()`
share the same random-duplication starting hypothesis and the same minimal
input shape, so they share a single dataset, `ratings_example`, rather than
one each. It is not called `binomial_example` so as not to mask
`stats::binomial`.

The package has two unrelated Hofmans models, so each keeps its domain in
both the function name and the dataset name: `hofmans_accumulation_example`
for `calc_hofmans_accumulation()` (one vehicle, several insertions) and
`hofmans_duplication_example` for `calc_hofmans_duplication()` (several
vehicles, one insertion each).

## MBD: the multivariate Beta-Binomial model

`calc_mbd()` implements Cheong's (2007) MBD model: observed (pairwise) or
imputed (three or more vehicles, via a Beta-Binomial fitted to the
subset's own mean audience and duplication) co-exposure probabilities are
combined via Waring's (1792) inclusion-exclusion theorem. Vehicles are
peeled off one at a time, in reverse aggregation order, expanding each
remaining pattern with a conditional Beta-Binomial allocation.

```{r}
data(mbd_cheong2007)
mbd_result <- do.call(calc_mbd, mbd_cheong2007)
mbd_result$reach
mbd_result$distribution
```

Unlike CSD/MSAD, Cheong documents that the model can produce negative
probabilities when combining four or more vehicles, because co-exposure for
three or more vehicles is imputed rather than observed. Cheong's own
four-vehicle example is corrected with a first-order consistency check;
the five-vehicle example, explicitly, is not -- Cheong identifies that a
higher-order check would be needed, which was never implemented, and
proposes (without developing it) Deming's iterative proportional fitting as
future work. `calc_mbd()` applies exactly the checks Cheong specifies
(first-order only) and, just like Cheong's own MBD-ADJ variant, zeroes out
any cell that remains negative and redistributes that mass proportionally --
exactly the same pattern already used by `calc_canex()`. This is reflected
in the diagnostics:

```{r}
mbd_result$diagnostics$negative_mass_adjusted
mbd_result$diagnostics$cells_adjusted
```

With four or more vehicles, `calc_mbd()` explicitly warns about this
limitation. Aggregation order can also alter the result (Cheong did not
investigate this systematically), and the model is capped at 12 vehicles due
to the exponential computational cost of the exposure grid -- the same
practical limit Cheong applied in the dissertation's own empirical
evaluation.

## CBD: the conditional Beta distribution

`calc_cbd()` implements the Conditional Beta Distribution (Leckenby & Kim,
reported in Kim, 1994; reviewed in Kim, 2005, pp. 59-64). Kim (2005) and
Cheong (2007) each independently rank CBD among the two or three most
accurate models they tested -- it is the strongest-recommended model this
package did not already have.

CBD shares its within-vehicle step exactly with `calc_mbd()`: each vehicle
is peeled off, in reverse aggregation order, from a (0,1) exposure state
into its own insertion-level Beta-Binomial distribution, via the same
conditional convolution (Danaher, 1992a). The two models differ only in how
the initial (0,1) joint grid across all vehicles is built. MBD imputes
co-exposure for three or more vehicles from a Beta-Binomial fit to each
subset's own mean audience and duplication, combined via Waring's
inclusion-exclusion. CBD instead applies Danaher's (1991) second-order
canonical expansion -- the same mechanism `calc_canex()`/`calc_csd()` use --
directly to every cell of the (0,1) grid, with no imputation step needed
for three or more vehicles.

```{r}
vehicles <- data.frame(
  insertions = c(2, 2, 2),
  R1 = c(0.4902, 0.0333, 0.0300),
  R2 = c(0.5805, 0.0502, 0.0371)
)
duplication <- matrix(
  c(NA, 0.0157, 0.0139,
    0.0157, NA, 0.0003,
    0.0139, 0.0003, NA),
  nrow = 3, byrow = TRUE
)
cbd_result <- calc_cbd(vehicles, duplication, aggregation_order = 1:3)
cbd_result$reach
cbd_result$distribution
```

These are the same three-vehicle inputs as `calc_csd()`'s Kim (2005)
example above, so the two can be compared directly: both use the same
between-vehicle canonical expansion, so their reach is close (though not
required to be identical, since they combine it with the within-vehicle
step differently). Kim (2005) reviews CBD's specification as an existing
model rather than walking through a numerical example of it the way she
does for her own CSD, so, unlike CSD, there is no published worked example
to validate this implementation against directly. It is validated instead
by construction: the (0,1) grid it builds recovers each vehicle's own R1 as
its exact marginal, and the full model reduces to the exact independent
convolution of the vehicles' own Beta-Binomial marginals when duplication
implies zero correlation -- both are checked in the test suite.

## Literature coverage and upcoming models

Kim's thesis does not offer the same level of reproducibility for every
model. `mediaPlanR`'s policy is not to add a new model until its primary
specification and a numeric reference that allows an independent regression
test are both available.

| Model | Accessible numeric evidence | v2 decision |
|---|---|---|
| CSD | Complete example, Kim's (2005) Tables 4.2.2.1-4.2.2.10 | Implemented and validated |
| MSAD | Algorithm given by Kim; detailed example referred to Lee (1988) | Implemented with invariants; external benchmark pending |
| CANEX | Danaher's primary formula and the CSD example that uses its reach | Implemented; keeps approximation diagnostics |
| BBD and Binomial | Closed-form formulas and marginals published by Kim | Implemented and tested at their limits |
| MBD | Complete specification and example (3 vehicles) from Cheong (2007); 4-5 vehicle examples show the model's own limits | Implemented, with a documented safety net and warnings for 4+ vehicles |
| CBD | Full specification, Kim (2005) pp. 59-64, reviewing Leckenby & Kim (Kim, 1994); ranked among the top models by both Kim (2005) and Cheong (2007) | Implemented; validated by construction (margin recovery, exact independence reduction) rather than an external numeric example |
| HBBD | Kim refers detailed steps to Boyd (1985) and Kim (1994) | Priority candidate; no equivalence claimed until those sources are obtained |
| DMD | Kim refers steps and examples to Boyd (1985) and Kim (1994); Aldas Manzano (1998) evaluated it empirically and ranked it below Metheringham/Kwerel geometric | Not planned: the one available empirical ranking recommends against it |
| HBD | MBD approximation described by Cheong (2007), with no verified numeric example of its own | Pending; would require the same level of validation as MBD |
| Multivariate NBD | Not Kim's example; requires Danaher's formulation for online processes | Future line, separate from the experimental univariate NBD |

Kim's Appendix B publishes observed and estimated distributions for 560
schedules and eight models. It is a valuable resource for a future external
benchmark suite, but it does not substitute for the primary specifications
of HBBD or DMD. Incorporating only the results columns without
reconstructing their inputs and algorithms would produce circular or
incomplete tests.

## References

- Agostini, J. M. (1961). How to estimate unduplicated audiences. *Journal
  of Advertising Research*, 1(3), 11-14.
- Aldas Manzano, J. (1998). *Modelos de determinacion de la cobertura y la
  distribucion de contactos en la planificacion de medios publicitarios
  impresos*. Doctoral dissertation, Universidad de Valencia.
- Chandon, J.-L. (1985). *A comparative study of media exposure models*.
  Doctoral dissertation, University of Pennsylvania.
- Cheong, Y. (2007). *Multivariate Beta Binomial Distribution Model as a Web
  Media Exposure Model*. Doctoral dissertation, The University of Texas at
  Austin.
- Danaher, P. J. (1991). A canonical expansion model for multivariate media
  exposure distributions. *Journal of Marketing Research*, 28(3), 361-367.
- Ehrenberg, A. S. C. (1959). The pattern of consumer purchases. *Journal of
  the Royal Statistical Society: Series C*, 8(1), 26-41.
- Kim, H. G. (2005). *A Canonical Sequential Aggregation Media Model*.
  Doctoral dissertation, The University of Texas at Austin.
- Leckenby, J. D., & Kishi, S. (1982). Performance of exposure distribution
  models. *Journal of Advertising Research*, 22(2), 35-44.
- Leckenby, J. D., & Rice, M. D. (1986). The declining reach phenomenon in
  exposure distribution models. *Journal of Advertising*, 15(3), 13-20.
- Metheringham, R. A. (1964). Measuring the net cumulative coverage of a
  print campaign. *Journal of Advertising Research*, 4(4), 23-28.
- Waring, E. (1792). *Meditationes Algebraicae*. Cambridge.

Lee & Burkart (1960) and Lee (1988), cited above for the net cumulative
audience Binomial model and for the detailed MSAD numerical example
respectively, are known to this package only at second hand, through Aldas
Manzano (1998) and Kim (2005); their own full bibliographic details have not
been independently verified and are therefore not listed as standalone
entries here.
