---
title: "Introduction to mediaPlanR"
author: "Manuel J. Sánchez-Franco"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to mediaPlanR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## What mediaPlanR provides

**mediaPlanR** brings together, in one package, the classical reach and
exposure-distribution models of advertising media planning and a validated
planning core: a plan object, plan metrics, model comparison, and
budget-constrained optimization of effective reach.

```{r setup}
library(mediaPlanR)
```

## The planning workflow

The planning functions start from a validated plan and keep every variable in
explicit units. Audience is people per insertion, cost is currency per
insertion, and population is the planning universe.

```{r plan}
plan <- media_plan(
  data.frame(
    channel = c("TV", "Radio", "Digital"),
    audience = c(300000, 180000, 120000),
    insertions = c(4, 6, 10),
    cost_per_insertion = c(18000, 3500, 1200),
    target_audience = c(180000, 90000, 84000)
  ),
  population = 1000000,
  target_audience = "target_audience"
)

metrics <- plan_metrics(plan)
reach <- estimate_reach(plan, model = "sainsbury")
reach
```

The returned distribution includes zero exposures. This makes it possible to
check that the probabilities sum to one, that reach equals `1 - P(X = 0)`, and
that average frequency equals impressions divided by people reached.

```{r invariants}
sum(reach$distribution$probability)
reach$reach$probability
reach$average_frequency
```

`estimate_reach()` is a plan-based front end for `calc_sainsbury()` and
`calc_binomial()`: it reads `audience`, `insertions` and `population` from the
plan and calls the requested model directly. `model = "sainsbury"` keeps each
vehicle's own audience (heterogeneous vehicles); `model = "binomial"` gives
every insertion the plan's insertion-weighted mean probability (homogeneous
vehicles). Both assume random duplication *and* random accumulation: a repeat
insertion in the same vehicle is treated as an independent trial, exactly like
an insertion in a different vehicle. This is a simplification, and it is the
hypothesis that the cited models state explicitly (Lee & Burkhart, 1960;
Chandon, 1976, as reviewed in Aldás Manzano, 1998, Sections 3.3.1.1 and
3.3.1.2, where it is cited as Chandon, 1985). It is also what allows them to
run without any observed duplication data. With one insertion per vehicle,
both reduce to their one-insertion-per-vehicle formulation (Aldás Manzano,
1998, Sections 3.2.2.1 and 3.2.2.2).

### Optimization with an optimality diagnostic

```{r optimization}
optimized <- optimize_media_plan(
  plan,
  budget = 60000,
  objective = "max_reach",
  effective_frequency = 2,
  max_insertions = c(4, 8, 12),
  method = "auto"
)

optimized$allocation
optimized$global_optimum
```

When the search space is manageable, every allocation is evaluated and
`global_optimum` is `TRUE`. In large spaces a greedy heuristic is used; it is
labeled as such and never presented as a global optimum. The heuristic adds,
at each step, between one and `effective_frequency` insertions to the channel
with the best gain in effective reach per unit of cost. Adding several
insertions at once matters because fewer insertions than the effective
frequency cannot by themselves create effective reach. Candidate allocations
routinely place several insertions in the same vehicle, so the optimizer offers
only the `sainsbury` and `binomial` reach models: the Negative-Binomial
approximation is scoped to continuous exposure processes, not to finite
insertion schedules.

## The family of reach and exposure-distribution models

All these models answer the same question: *what proportion of the population
is reached, and how many times, by a given media plan?* They differ in the
**shape of plan** they were derived for, and in their assumptions about the
**duplication** of audiences across vehicles and about the **heterogeneity** of
individual exposure probabilities. Plan shape follows Aldás Manzano's (1998)
three-way split, which the package keeps visible in its function names.
Accumulation models (Beta-Binomial, Hofmans accumulation) are derived for one
vehicle with several insertions. Duplication models (Agostini, Hofmans
duplication) are derived for several vehicles with one insertion each. Net
cumulative audience models cover several vehicles with several insertions.
Sainsbury and Binomial cover both simple cases because each is one model at two
levels of generality (see the `insertions` argument).

| Model | Function | Plan shape | Duplication | Heterogeneity |
|---|---|---|---|---|
| Sainsbury | `calc_sainsbury()` / `estimate_reach(model = "sainsbury")` | 1 or more insertions, m vehicles | Random (independence and accumulation) | Vehicles only |
| Binomial | `calc_binomial()` / `estimate_reach(model = "binomial")` | 1 or more insertions, m vehicles | Random (independence and accumulation) | None (average vehicle) |
| Agostini (duplication) | `calc_agostini_duplication()` | 1 insertion, m vehicles | Observed pairwise, one coefficient `k` | Reach only |
| Hofmans (duplication) | `calc_hofmans_duplication()` | 1 insertion, m vehicles | Observed pairwise, pairwise coefficient | Reach only |
| Beta-Binomial | `calc_beta_binomial()` | 1 vehicle, n insertions | Implicit (fitted to `R1`, `R2`) | Individuals (Beta-Binomial) |
| Hofmans (accumulation) | `calc_hofmans_accumulation()` | 1 vehicle, n insertions | Implicit (fitted to `R1`, `R2`, optionally `R3`) | Reach only |
| Metheringham | `calc_metheringham()` | n insertions, m vehicles | Observed, averaged over insertion pairs | Individuals (Beta-Binomial) |
| CANEX | `calc_canex()` | n insertions, m vehicles | Observed, canonical correlations | Per vehicle (Beta-Binomial) |
| CBD (conditional Beta) | `calc_cbd()` | n insertions, m vehicles | Observed; canonical expansion at the (0,1) level | Per vehicle (Beta-Binomial) |
| Beta-Binomial fitted to a reach | `fit_bbd_to_reach()` | n insertions, m vehicles | Reach supplied as a constraint | Individuals (Beta-Binomial) |
| CSD (canonical sequential aggregation) | `calc_csd()` | n insertions, m vehicles | Observed; canonical reach per sub-schedule | Per vehicle (Beta-Binomial) |
| MSAD (Morgensztern) | `calc_msad()` | n insertions, m vehicles | Observed; Morgensztern reach per sub-schedule | Per vehicle (Beta-Binomial) |
| MBD (multivariate Beta-Binomial) | `calc_mbd()` | n insertions, m vehicles | Observed pairs; imputed for 3+ vehicles (Waring's theorem) | Per vehicle (Beta-Binomial) |
| Negative-Binomial approximation | `nbd_exposure_distribution()` | Continuous process, not a finite schedule | None across vehicles | Poisson-Gamma, unbounded |

### Example: three vehicles, one insertion each, different assumptions

Sainsbury and Binomial need only the audiences. Agostini and Hofmans also use
the audience duplicated between each pair of vehicles, and Metheringham uses
the average duplication.

```{r duplication}
data(ratings_example)
data(duplication_example)

reach_sainsbury <- do.call(calc_sainsbury, ratings_example)
reach_binomial <- do.call(calc_binomial, ratings_example)
reach_agostini <- do.call(calc_agostini_duplication, duplication_example)
reach_hofmans <- do.call(calc_hofmans_duplication, duplication_example)
reach_metheringham <- calc_metheringham(
  audiences = duplication_example$audiences,
  insertions = c(1, 1, 1),
  duplication_matrix = duplication_example$duplication_matrix,
  population = duplication_example$population
)

data.frame(
  model = c("Sainsbury", "Binomial", "Agostini (k = 1.125)", "Hofmans (duplication)",
            "Metheringham"),
  reach_pct = round(c(
    reach_sainsbury$reach$percent,
    reach_binomial$reach$percent,
    reach_agostini$reach$percent,
    reach_hofmans$reach$percent,
    reach_metheringham$reach$percent
  ), 2)
)
```

The observed duplication in `duplication_example` is higher than random
duplication would produce for every pair of vehicles, so the models that use it
estimate a lower reach than Sainsbury and Binomial. Agostini's formula applies
one empirical coefficient (`k = 1.125`, Agostini's own estimate) to the total
duplication, whereas Hofmans' formula computes a coefficient from each pair's
own audiences, so no coefficient has to be fitted. Both are ad hoc formulas: they
return reach only, and `calc_agostini_duplication()` and
`calc_hofmans_duplication()` warn when the reach they compute leaves its logical
range.

### Several insertions per vehicle: Metheringham

`calc_metheringham()` averages audiences and duplications over all pairs of
insertions, both between vehicles and within a vehicle, to define an "average"
vehicle, and fits a Beta-Binomial to the plan's `N = sum(insertions)`
insertions (Aldás Manzano, 1998, Section 3.3.1.5). The diagonal of the
duplication matrix holds the audience duplicated between two insertions in the
same vehicle.

```{r metheringham}
data(metheringham_example)
metheringham <- do.call(calc_metheringham, metheringham_example)
metheringham$reach$percent
length(metheringham$distribution$percent)   # zero to N = 12 exposures

# The mean of the distribution equals the plan's gross exposures per person
sum(seq_along(metheringham$distribution$percent) *
      metheringham$distribution$percent) / 100
with(metheringham_example, sum(audiences * insertions) / population)
```

### One vehicle, several insertions: Beta-Binomial and Hofmans

The Beta-Binomial accumulation model needs the audience after one and two
insertions. Hofmans' accumulation model, with a constant coefficient, needs the
same two figures; an observed third insertion (`R3`) lets the coefficient vary
with the number of insertions.

```{r accumulation}
beta_binomial <- calc_beta_binomial(A1 = 500000, A2 = 550000, P = 1000000, n = 5)
beta_binomial$reach$percent

hofmans <- calc_hofmans_accumulation(R1 = 0.06, R2 = 0.103, N = 5)
hofmans$results
```

## Beta-Binomial versus Negative-Binomial: finite opportunities and count processes

Both `calc_beta_binomial()` and the Negative-Binomial distribution assume that
each individual has a personal exposure probability (or rate) that varies
across the population, but they mix different distributions. The Beta-Binomial
mixes a Binomial with a fixed number of insertions with a Beta; the
Negative-Binomial mixes a Poisson with a Gamma (Ehrenberg, 1959). The
Negative-Binomial has unbounded support, so it suits page views, ad-server
exposures and other processes with continuous opportunities. It should not be
used to suggest that a finite-insertion schedule can register more exposures
than insertions.

When individual counts are observed, `fit_nbd_exposure()` estimates the mean and
the heterogeneity by maximum likelihood. For scenarios with external
parameters, `nbd_exposure_distribution()` keeps the last cell as an open tail
and never presents it as an exact count. Danaher (2007) develops the
multivariate Negative-Binomial model that represents dependence across
websites; the functions here are deliberately univariate.

```{r nbd}
counts <- c(rep(0, 40), rep(1, 25), rep(2, 15), rep(3, 8), 5, 7)
nbd_fit <- fit_nbd_exposure(counts)
nbd_fit$estimates

nbd_scenario <- nbd_exposure_distribution(
  mean_contacts = 2.5, size = 1.7, report_max = 9, opportunities = 9
)
nbd_scenario$diagnostics$probability_above_opportunities
```

The last diagnostic is the probability mass that the unbounded model assigns to
more exposures than the finite opportunities allow.

## CANEX: heterogeneity and jointly observed duplication

`calc_canex()` implements Danaher's (1991) canonical expansion model. It combines
a Beta-Binomial distribution for each vehicle with a correlation term that
adjusts the joint probability according to the duplication observed between
each pair of vehicles.

```{r canex}
data(canex_example)
canex_result <- do.call(calc_canex, canex_example)
canex_result
canex_result$diagnostics
```

The number of exposure combinations grows exponentially with the number of
vehicles and insertions, so the model is intended for small and medium
schedules. For very large schedules `calc_canex()` stops with an informative
error instead of exhausting memory. Because the second-order expansion can
produce negative probabilities, `calc_canex()` sets them to zero and
renormalizes; the diagnostics report how much mass was truncated.

## CSD and MSAD: two reach estimators, one sequential architecture

`calc_csd()` reproduces the complete numerical example of Kim (2005): three
vehicles, Beta-Binomial marginals, second-order canonical-expansion reach, and
non-random convolution that preserves the margins. `csd_kim2005` contains the
published inputs (reach and duplication figures, with attribution).

```{r csd}
data(csd_kim2005)
csd_result <- do.call(calc_csd, csd_kim2005)
csd_result$reach
csd_result$distribution
csd_result$steps
```

The exact results can differ by a few hundredths of a percentage point from
Kim's tables, because the dissertation rounds parameters and intermediate
matrices to four decimals; the function keeps the precision of its inputs.

### MSAD: Morgensztern's sequential aggregation

`calc_msad()` implements the structure described by Leckenby and Rice (1986)
and Kim (2005): a Beta-Binomial marginal for each vehicle, the Morgensztern
reach for each sub-schedule, and a non-random convolution that preserves the
marginals and sets the zero-exposure probability to `1 - reach`.

```{r msad}
data(msad_kim2005)
msad_result <- do.call(calc_msad, msad_kim2005)
msad_result$reach
msad_result$steps
msad_result$diagnostics
```

By default the vehicles with the highest audience are aggregated first, as Kim
indicates. An explicit permutation tests the sensitivity to the order. The
function stops if the duplications or the Morgensztern reach violate
probabilistic bounds.

Kim specifies MSAD but refers to Lee (1988) for the complete numerical example.
`msad_kim2005` therefore reuses the inputs of Kim's CSD example, and its output
is a derived calculation, not an MSAD result published in the dissertation.

## Validation against observed data

Observed data are always supplied by the analyst. For one schedule they must
be a table with the columns `contacts` and `observed`, explicitly including zero
exposures, and the scale of the values (counts, probabilities or percentages)
must be declared.

```{r evaluation}
kim_observed <- data.frame(
  contacts = 0:6,
  observed = c(38.41, 17.89, 39.66, 2.67, 1.36, 0, 0)
)

csd_evaluation <- evaluate_exposure_model(
  observed = kim_observed,
  predicted = csd_result,
  observed_scale = "percent"
)

csd_evaluation$summary
csd_evaluation$by_schedule
```

If the prediction is also a table, its columns must be named `contacts` and
`predicted`, and `predicted_scale` must be declared. To evaluate several
schedules at once, add an identifier column to both tables and name it in
`schedule_col`.

The evaluator requires identical supports. It does not interpret an open tail
of the Negative-Binomial approximation as an exact frequency: observed and
predicted tails must first be collapsed at the same threshold. Besides Kim's
average error in reach and in the exposure distribution, it returns absolute
reach error, total variation, cell mean absolute error and mean-exposure bias.

## MBD: the multivariate Beta-Binomial model

`calc_mbd()` implements Cheong's (2007) model. Pairwise co-exposure
probabilities are observed; those of three or more vehicles are imputed from a
Beta-Binomial fitted to the subset's own mean audience and duplication. They are
combined into the joint zero/one grid by Waring's (1792) inclusion-exclusion
theorem. Vehicles are then peeled off one at a time, in reverse aggregation
order, expanding each remaining pattern with a conditional Beta-Binomial
allocation.

```{r mbd}
data(mbd_cheong2007)
mbd_result <- do.call(calc_mbd, mbd_cheong2007)
mbd_result$reach
mbd_result$distribution
```

Unlike CSD and MSAD, the model can produce negative probabilities when it
combines four or more vehicles, because the co-exposure of three or more
vehicles is imputed rather than observed. Cheong's four-vehicle example is
corrected by a first-order consistency check. The five-vehicle example is not:
Cheong notes that a higher-order check would be needed, which was never
implemented, and proposes iterative proportional fitting (Deming, 1943) as
future work. `calc_mbd()` applies exactly the checks Cheong specifies (first
order only) and, like Cheong's MBD-ADJ variant, sets any cell that remains
negative to zero and redistributes that mass proportionally. The diagnostics
report it:

```{r mbd-diagnostics}
mbd_result$diagnostics$negative_mass_adjusted
mbd_result$diagnostics$cells_adjusted
```

With four or more vehicles `calc_mbd()` warns about this limitation. The
aggregation order can also alter the result (Cheong did not investigate this
systematically), and the model is limited to 12 vehicles because of the
exponential cost of the exposure grid, the same practical limit Cheong applied
in the dissertation's own evaluation.

## CBD: the conditional Beta distribution

`calc_cbd()` implements the Conditional Beta Distribution of Leckenby and Kim,
reported in Kim (1994) and reviewed in Kim (2005, pp. 59-64). In Kim's (2005)
evaluation CBD is among the three most accurate models, together with CSD and
CANEX; in Cheong's (2007) it is second only to CANEX for the complete exposure
distribution.

CBD shares its within-vehicle step with `calc_mbd()`. Each vehicle is peeled off,
in reverse aggregation order, from a (0,1) exposure state into its own
insertion-level Beta-Binomial distribution by the same conditional convolution
(Danaher, 1992a). The two models differ only in how the initial (0,1) joint
grid across all vehicles is built. MBD imputes the co-exposure of three or more
vehicles from a Beta-Binomial fitted to each subset's own mean audience and
duplication and combines it with Waring's inclusion-exclusion theorem. CBD
instead applies Danaher's (1991) second-order canonical expansion, the mechanism
used by `calc_canex()` and `calc_csd()`, directly to every cell of the (0,1)
grid, with no imputation step.

```{r cbd}
cbd_result <- do.call(calc_cbd, csd_kim2005)
cbd_result$reach
cbd_result$distribution
```

These are the same three-vehicle inputs as the CSD example, so the two can be
compared directly. Both use the same between-vehicle canonical expansion, so
their reach is close, although it is not required to be identical because the
within-vehicle step differs. Kim (2005) reviews the specification of CBD as an
existing model and does not present a numerical example of it, so there is no
published worked example to validate this implementation against. It is
validated by construction: the (0,1) grid recovers each vehicle's own `R1` as its
exact marginal, and the model reduces to the exact convolution of the vehicles'
own Beta-Binomial marginals when the duplications imply zero correlation. Both
properties are checked in the test suite.

## Example datasets

Models with complex inputs come with datasets whose elements match the
function's arguments, so they can be used with `do.call()`:

```{r datasets}
data(canex_example)
str(canex_example)
do.call(calc_canex, canex_example)$reach$percent
```

Datasets whose name ends in `_example` are original illustrative scenarios,
meant for quick experimentation. `csd_kim2005`, `msad_kim2005` and
`mbd_cheong2007` instead reproduce the minimal numeric inputs (reach and
duplication figures) published by Kim (2005) and Cheong (2007), so that anyone
can verify that `calc_csd()`, `calc_msad()` and `calc_mbd()` reproduce the
published examples. They are bare factual figures reused for validation, not a
reproduction of the dissertations. `ratings_example` serves `calc_sainsbury()`
and `calc_binomial()`, and `duplication_example` serves
`calc_agostini_duplication()` and `calc_hofmans_duplication()`, because each
pair shares the same input shape. No dataset is named `binomial`, so that
`stats::binomial` is never masked.

The package has two unrelated Hofmans models, so both function names carry their
domain: `calc_hofmans_accumulation()` (one vehicle, several insertions, with
`hofmans_accumulation_example`) and `calc_hofmans_duplication()` (several
vehicles, one insertion each, with `duplication_example`).

## Scope

The package adds a model only when its primary specification and a numeric
reference that allows an independent regression test are available.

| Model | Numeric evidence | Status |
|---|---|---|
| CSD | Complete example, Kim (2005), Tables 4.2.2.1-4.2.2.10 | Implemented and validated |
| MSAD | Algorithm given by Kim; detailed example referred to Lee (1988) | Implemented; external benchmark pending |
| CANEX | Danaher's (1991) formula and the CSD example that uses its reach | Implemented; reports approximation diagnostics |
| Beta-Binomial and Binomial | Closed-form formulas | Implemented and tested at their limits |
| MBD | Complete specification and three-vehicle example, Cheong (2007); four- and five-vehicle examples show the model's limits | Implemented, with a documented safety net and warnings |
| CBD | Specification in Kim (2005, pp. 59-64) after Kim (1994) | Implemented; validated by construction |
| HBBD (Hofmans Beta-Binomial) | Kim refers detailed steps to Boyd (1985) and Kim (1994) | Not implemented: primary steps not available |
| DMD (Dirichlet multinomial) | Kim refers steps and examples to Boyd (1985) and Kim (1994); Aldás Manzano (1998) ranked it below the Metheringham and Kwerel geometric models | Not implemented |
| HBD | Approximation of MBD described by Cheong (2007), without a verified numeric example | Not implemented |
| Multivariate Negative-Binomial | Requires Danaher's (2007) formulation for online processes | Not implemented |

## References

- Agostini, J. M. (1961). How to estimate unduplicated audiences. *Journal of
  Advertising Research*, 1(3), 11-14. doi:10.1080/00218499.1961.12519620
- Aldás Manzano, J. (1998). *Modelos de determinación de la cobertura y la
  distribución de contactos en la planificación de medios publicitarios
  impresos* (Models for determining reach and exposure distribution in print
  media planning). Doctoral dissertation, Universidad de Valencia, Spain.
- Boyd, M. M. (1985). *The Dirichlet multivariate multinomial distribution as a
  magazine exposure model*. Unpublished doctoral dissertation, University of
  Illinois at Urbana-Champaign.
- Chandon, J.-L. (1976). *A comparative study of media exposure models*.
  Unpublished doctoral dissertation, Northwestern University, Evanston, IL.
- Cheong, Y. (2007). *Multivariate Beta Binomial Distribution Model as a Web
  Media Exposure Model*. Doctoral dissertation, The University of Texas at
  Austin.
- Danaher, P. J. (1991). A canonical expansion model for multivariate media
  exposure distributions: A generalization of the "duplication of viewing law".
  *Journal of Marketing Research*, 28(3), 361-367.
  doi:10.1177/002224379102800311
- Danaher, P. J. (1992a). A Markov-chain model for multivariate
  magazine-exposure distributions. *Journal of Business & Economic Statistics*,
  10(4), 401-407. doi:10.1080/07350015.1992.10509915
- Danaher, P. J. (2007). Modeling page views across multiple websites with an
  application to Internet reach and frequency prediction. *Marketing Science*,
  26(3), 422-437. doi:10.1287/mksc.1060.0226
- Deming, W. E. (1943). *Statistical adjustment of data*. New York: Dover.
- Ehrenberg, A. S. C. (1959). The pattern of consumer purchases. *Applied
  Statistics*, 8(1), 26-41. doi:10.2307/2985810
- Hofmans, P. (1966). Measuring the cumulative net coverage of any combination
  of media. *Journal of Marketing Research*, 3(3), 269-278.
  doi:10.1177/002224376600300307
- Kim, H. (1994). *A conditional beta distribution model for advertising
  reach/frequency estimation*. Unpublished doctoral dissertation, The University
  of Texas at Austin.
- Kim, H. G. (2005). *A Canonical Sequential Aggregation Media Model*. Doctoral
  dissertation, The University of Texas at Austin.
- Leckenby, J. D., & Rice, M. D. (1986). The declining reach phenomenon in
  exposure distribution models. *Journal of Advertising*, 15(3), 13-20.
  doi:10.1080/00913367.1986.10673014
- Lee, A. M., & Burkhart, A. J. (1960). Some optimization problems in advertising
  media planning. *Operational Research Quarterly*, 11(3), 113-122.
- Lee, H.-K. (1988). *Sequential aggregation advertising media models*.
  Unpublished doctoral dissertation, The University of Texas at Austin.
- Metheringham, R. A. (1964). Measuring the net cumulative coverage of a print
  campaign. *Journal of Advertising Research*, 4(4), 23-28.
  doi:10.1080/00218499.1964.12519751
- Waring, E. (1792). *Meditationes Algebraicae*. Cambridge.

Morgensztern (1970), whose reach formula underlies MSAD, is known to this
package through Aldás Manzano (1998) and Kim (2005). Lee and Burkhart (1960) and
Lee (1988) are cited as they appear in Kim (2005). The bibliographic details of
Chandon (1976), Boyd (1985), Kim (1994) and Lee (1988) are those given in the
reference list of Kim (2005).
