---
title: "Introduction to mediaPlanR"
author: "Manuel J. Sánchez-Franco"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to mediaPlanR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## What mediaPlanR provides

**mediaPlanR** brings together, in a single R package, the classical reach
and contact-distribution models used in advertising media planning, along
with a validated planning core (plan, metrics, model comparison, and
budget-constrained reach optimization).

```{r setup}
library(mediaPlanR)
```

## The version 2 workflow

The main API starts from a validated plan and preserves each variable's
units. Audience represents people per insertion; cost represents currency
per insertion; population is the planning universe.

```{r v2-plan}
plan <- media_plan(
  data.frame(
    channel = c("TV", "Radio", "Digital"),
    audience = c(300000, 180000, 120000),
    insertions = c(4, 6, 10),
    cost_per_insertion = c(18000, 3500, 1200),
    target_audience = c(180000, 90000, 84000)
  ),
  population = 1000000,
  target_audience = "target_audience"
)

metrics <- plan_metrics(plan)
reach <- estimate_reach(plan, model = "independent")
reach
```

The returned distribution includes zero contacts. This lets you check that
the probabilities sum to one, that reach equals `1 - P(X=0)`, and that
average frequency matches impressions divided by people reached.

```{r v2-invariants}
sum(reach$distribution$probability)
reach$reach$probability
reach$average_frequency
```

### Optimization with an optimality diagnostic

```{r v2-optimization}
optimized <- optimize_media_plan(
  plan,
  budget = 60000,
  objective = "max_reach",
  effective_frequency = 2,
  max_insertions = c(4, 8, 12),
  method = "auto"
)

optimized$allocation
optimized$global_optimum
```

When the search space is manageable it is evaluated exhaustively and
`global_optimum = TRUE`. In large spaces, the heuristic mode is identified as
such; it is never presented as a global optimum.

## The family of reach and contact-distribution models

All of these models answer the same question -- *what proportion of the
target population is reached, and how many times, by a given media plan?*
-- but they differ in the assumptions they make about the **duplication** of
audiences across vehicles and about the **heterogeneity** of individual
exposure probabilities.

| Model | Function | Assumed duplication | Heterogeneity |
|---|---|---|---|
| Sainsbury | `calc_sainsbury()` | Random (independence) | Yes, across vehicles |
| Binomial | `calc_binomial()` | Random (independence) | No (average vehicle) |
| Beta-Binomial | `calc_beta_binomial()` | Implicit (one vehicle, n insertions) | Yes, across individuals (Binomial-Beta mixture) |
| Experimental NBD | `nbd_exposure_distribution()` | Does not represent cross-vehicle dependence | Yes, Poisson-Gamma heterogeneity in an unbounded process |
| Metheringham | `calc_metheringham()` | Observed (averaged) | Yes, via a Beta distribution |
| Hofmans | `calc_hofmans()` | Observed (R1, R2) | Non-linear empirical curve |
| Agostini | `calc_agostini()` | Corrected with a k coefficient | Random + empirical correction |
| BBD fitted to external reach | `fit_bbd_to_reach()` | Reach is supplied as a constraint | Yes, via Beta-Binomial |
| CSD (canonical sequential aggregation) | `calc_csd()` | Observed; canonical reach per sub-schedule with preserved margins | Yes, per-vehicle BBD marginal |
| MSAD (Morgensztern) | `calc_msad()` | Observed and preserved during sequential aggregation | Yes, per-vehicle BBD marginal |
| CANEX | `calc_canex()` | Observed, correlation matrix | Yes, per vehicle (Beta-Binomial) |

### Example: three vehicles, the same plan, different assumptions

```{r}
audiences <- c(300000, 400000, 200000)
population <- 1000000

reach_sainsbury <- calc_sainsbury(audiences, population)
reach_binomial <- calc_binomial(audiences, population)
reach_agostini_independence <- calc_agostini(audiences, population, k = 1)
reach_agostini_corrected <- calc_agostini(audiences, population, k = 0.9)

data.frame(
  model = c("Sainsbury", "Binomial", "Agostini (k=1)", "Agostini (k=0.9)"),
  reach_pct = round(c(
    reach_sainsbury$reach$percent,
    reach_binomial$reach$percent,
    reach_agostini_independence$reach$percent,
    reach_agostini_corrected$reach$percent
  ), 2)
)
```

Note that, with `k = 1`, `calc_agostini()` matches `calc_sainsbury()`
exactly: both start from the same random-duplication assumption, applied
iteratively in Agostini's case. With `k = 0.9`, predicted duplication is
reduced and estimated reach therefore increases relative to the pure
independence assumption.

## Beta-Binomial vs. NBD: finite opportunities and count processes

Both `calc_beta_binomial()` and the NBD assume each individual has their own
personal exposure probability (or rate), heterogeneous across the
population, but they differ in their distributional assumption: the former
mixes a Binomial (n fixed insertions) with a Beta; the NBD mixes a Poisson
with a Gamma. That difference is substantive: the NBD has unbounded support
and is appropriate for page views, ad-server contacts, or other processes
with continuous opportunities -- not for pretending that a traditional
schedule allows more contacts than insertions.

When individual observed counts are available, `fit_nbd_exposure()`
estimates the mean and heterogeneity by maximum likelihood. For scenarios
with external parameters, `nbd_exposure_distribution()` keeps the last cell
as an open tail and never presents it as an exact count:

```{r}
counts <- c(rep(0, 40), rep(1, 25), rep(2, 15), rep(3, 8), 5, 7)
nbd_fit <- fit_nbd_exposure(counts)
nbd_fit$estimates

nbd_scenario <- nbd_exposure_distribution(
  mean_contacts = 2.5, size = 1.7, report_max = 9, opportunities = 9
)
nbd_scenario$diagnostics$probability_above_opportunities
```

That last diagnostic quantifies the mass an unbounded NBD assigns above the
finite opportunities.

## CSD and MSAD: two reach estimators, one sequential architecture

`calc_csd()` reproduces Kim's (2005) complete numerical example: three
vehicles, BBD marginals, second-order canonical-expansion reach, and
non-random convolution with margin preservation.

```{r}
data(csd_kim2005)
csd_result <- do.call(calc_csd, csd_kim2005)
csd_result$reach
csd_result$distribution
csd_result$steps
```

The exact results may differ by a few hundredths of a percentage point from
Kim's tables because the thesis rounds parameters and intermediate matrices
to four decimals. The function preserves the precision of its inputs.

### MSAD: Morgensztern's sequential aggregation

`calc_msad()` implements the structure described by Leckenby and Rice (1986)
and Kim (2005): a per-vehicle BBD marginal, Morgensztern reach for each
sub-schedule, and non-random convolution that preserves the marginals and
sets the zero-contact probability to `1 - reach`.

```{r}
data(msad_kim2005)
msad_result <- do.call(calc_msad, msad_kim2005)
msad_result$reach
msad_result$steps
msad_result$diagnostics
```

The default rule aggregates the highest-audience vehicles first, as Kim
indicates. An explicit permutation can also be supplied to test sensitivity
to order. The function stops if the duplications or the Morgensztern reach
violate probabilistic bounds.

Kim specifies MSAD but refers to Lee (1988) for the complete numerical
example. `msad_kim2005` therefore reuses the inputs from Kim's CSD example
and labels its output as a derived calculation, not an MSAD result published
in the thesis.

## Validation against observed data

Observed data are always supplied by the analyst. For one schedule, they
must be provided as a table with the columns `contacts` and `observed`,
explicitly including zero contacts. Whether the values are counts,
probabilities, or percentages must also be declared:

```{r}
kim_observed <- data.frame(
  contacts = 0:6,
  observed = c(38.41, 17.89, 39.66, 2.67, 1.36, 0, 0)
)

csd_evaluation <- evaluate_exposure_model(
  observed = kim_observed,
  predicted = csd_result,
  observed_scale = "percent"
)

csd_evaluation$summary
csd_evaluation$by_schedule
```

If the prediction is also supplied as a table, its columns must be named
`contacts` and `predicted`, and `predicted_scale` must be declared. To
evaluate several schedules at once, add an identifier column to both tables
and name it via `schedule_col`.

The evaluator requires identical supports. It does not interpret an NBD open
tail as an exact frequency: before evaluation, the observed and predicted
tails must be collapsed at the same threshold. Besides Kim's historical
AER/APE, it also returns absolute reach error, total variation, cell MAE,
and mean-contact bias.

## The CANEX model: heterogeneity and jointly observed duplication

`calc_canex()` implements Danaher's (1991) canonical expansion model, which
combines a per-vehicle Beta-Binomial distribution with a correlation term
that adjusts the joint probability according to the observed duplication
between each pair of vehicles:

```{r}
vehicles <- data.frame(
  k = c(2, 2),
  R1 = c(0.4902, 0.033),
  R2 = c(0.5805, 0.0502)
)
duplications <- matrix(
  c(1.000, 0.0157,
    0.0157, 1.000),
  nrow = 2, byrow = TRUE
)

canex_result <- calc_canex(vehicles, duplications, population = 1000000)
canex_result
```

The number of exposure combinations to evaluate grows exponentially with the
number of vehicles and insertions, so the model is intended for small- or
medium-sized schedules; for very large schedules, `calc_canex()` stops with
an informative error instead of exhausting available memory.

## Ready-to-use example datasets

Models with more complex input structures have datasets with descriptive
names, whose elements match the function's arguments:

```{r}
data(canex_example)
str(canex_example)
do.call(calc_canex, canex_example)$total_reach
```

For CSD, MSAD, and MBD there are two example datasets per model.
`csd_example`, `msad_example`, and `mbd_example` are original, illustrative,
self-contained scenarios meant for quick experimentation with `do.call()`.
`csd_kim2005`, `msad_kim2005`, and `mbd_cheong2007` instead reproduce the
minimal numeric values (audiences and duplication figures) published by Kim
(2005) and Cheong (2007) in their doctoral dissertations, included solely so
anyone can verify that `calc_csd()`, `calc_msad()`, and `calc_mbd()`
reproduce their respective published examples: they are bare factual data
points reused for validation, not a creative or substantial reproduction of
those dissertations. `bbd_reach_example` plays the same illustrative role
for `fit_bbd_to_reach()`.
The dataset for `calc_binomial()` is called `binomial_plan` so as not to
mask `stats::binomial`.

## MBD: the multivariate Beta-Binomial model

`calc_mbd()` implements Cheong's (2007) MBD model: observed (pairwise) or
imputed (three or more vehicles, via a Beta-Binomial fitted to the
subset's own mean audience and duplication) co-exposure probabilities are
combined via Waring's (1792) inclusion-exclusion theorem. Vehicles are
peeled off one at a time, in reverse aggregation order, expanding each
remaining pattern with a conditional Beta-Binomial allocation.

```{r}
data(mbd_cheong2007)
mbd_result <- do.call(calc_mbd, mbd_cheong2007)
mbd_result$reach
mbd_result$distribution
```

Unlike CSD/MSAD, Cheong documents that the model can produce negative
probabilities when combining four or more vehicles, because co-exposure for
three or more vehicles is imputed rather than observed. Her own
four-vehicle example is corrected with a first-order consistency check; her
five-vehicle example, explicitly, is not -- Cheong identifies that a
higher-order check would be needed, which she never implemented, and
proposes (without developing it) Deming's iterative proportional fitting as
future work. `calc_mbd()` applies exactly the checks Cheong specifies
(first-order only) and, just like her own MBD-ADJ variant, zeroes out any
cell that remains negative and redistributes that mass proportionally --
exactly the same pattern already used by `calc_canex()`. This is reflected
in the diagnostics:

```{r}
mbd_result$diagnostics$negative_mass_adjusted
mbd_result$diagnostics$cells_adjusted
```

With four or more vehicles, `calc_mbd()` explicitly warns about this
limitation. Aggregation order can also alter the result (Cheong did not
investigate this systematically), and the model is capped at 12 vehicles due
to the exponential computational cost of the exposure grid -- the same
practical limit Cheong applied in her own empirical evaluation.

## Literature coverage and upcoming models

Kim's thesis does not offer the same level of reproducibility for every
model. `mediaPlanR`'s policy is not to add a new model until its primary
specification and a numeric reference that allows an independent regression
test are both available.

| Model | Accessible numeric evidence | v2 decision |
|---|---|---|
| CSD | Complete example, Kim's (2005) Tables 4.2.2.1-4.2.2.10 | Implemented and validated |
| MSAD | Algorithm given by Kim; detailed example referred to Lee (1988) | Implemented with invariants; external benchmark pending |
| CANEX | Danaher's primary formula and the CSD example that uses its reach | Implemented; keeps approximation diagnostics |
| BBD and Binomial | Closed-form formulas and marginals published by Kim | Implemented and tested at their limits |
| MBD | Complete specification and example (3 vehicles) from Cheong (2007); 4-5 vehicle examples show the model's own limits | Implemented, with a documented safety net and warnings for 4+ vehicles |
| HBBD | Kim refers detailed steps to Boyd (1985) and Kim (1994) | Priority candidate; no equivalence claimed until those sources are obtained |
| CBD | Kim refers the numeric example to Kim (1994) | Pending the primary source |
| DMD | Kim refers steps and examples to Boyd (1985) and Kim (1994) | Pending; requires validating the full multivariate matrix |
| HBD | MBD approximation described by Cheong (2007), with no verified numeric example of its own | Pending; would require the same level of validation as MBD |
| Multivariate NBD | Not Kim's example; requires Danaher's formulation for online processes | Future line, separate from the experimental univariate NBD |

Kim's Appendix B publishes observed and estimated distributions for 560
schedules and eight models. It is a valuable resource for a future external
benchmark suite, but it does not substitute for the primary specifications
of HBBD, CBD, or DMD. Incorporating only the results columns without
reconstructing their inputs and algorithms would produce circular or
incomplete tests.

## References

- Aldas Manzano, J. (1998). *Modelos de determinacion de la cobertura y la
  distribucion de contactos en la planificacion de medios publicitarios
  impresos*. Doctoral dissertation, Universidad de Valencia.
- Agostini, J. M. (1961). How to estimate unduplicated audiences. *Journal
  of Advertising Research*, 1(3), 11-14.
- Danaher, P. J. (1991). A canonical expansion model for multivariate media
  exposure distributions. *Journal of Marketing Research*, 28(3), 361-367.
- Kim, H. G. (2005). *A Canonical Sequential Aggregation Media Model*.
  Doctoral dissertation, The University of Texas at Austin.
- Ehrenberg, A. S. C. (1959). The pattern of consumer purchases. *Journal of
  the Royal Statistical Society: Series C*, 8(1), 26-41.
- Leckenby, J. D., & Kishi, S. (1982). Performance of exposure distribution
  models. *Journal of Advertising Research*, 22(2), 35-44.
- Metheringham, R. A. (1964). Measuring the net cumulative coverage of a
  print campaign. *Journal of Advertising Research*, 4(4), 23-28.
